---
title: "ImmuneSignatures2: Reproduction of analysis and figures for pre-vaccination manuscript"
output: 
  html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  dataDir: "/share/files/HIPC/IS2/@files/data"
  dataCacheDir: "/share/files/HIPC/IS2/@files/cache/prevax"
  cache.path: "/share/files/HIPC/IS2/@files/cache/prevax/report_cache"
  cache: false
---

<script>
setTimeout(() => {
  btns = document.querySelectorAll(".code-folding-btn");
  btns.forEach((x) => (x.style.display = "none"));
  console.log("hid the buttons");
}, 2000);
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      message    = FALSE,
                      warning    = FALSE, 
                      cache = params$cache,
                      cache.path = params$cache.path)
dataDir <- params$dataDir
dataCacheDir <- params$dataCacheDir
```

```{r load-packages}
suppressPackageStartupMessages({
  library(package = "knitr")
  library(package = "Biobase")
  library(package = "RColorBrewer")
  library(package = "cluster")
  library(package = "ComplexHeatmap")
  library(package = "ggbeeswarm")
  library(package = "circlize")
  library(package = "glue")
  library(package = "boot")
  library(package = "pvca")
  library(package = "lme4")
  library(package = "biomaRt")
  library(package = "ImmuneSignatures2")
  library(package = "pROC")
  library(package = "parallel")
  library(package = "MetaIntegrator")
  library(package = "AnnotationDbi")
  library(package = "Seurat")
  library(package = "magrittr")
  library(package = "ggrepel")
  library(package = "colorspace")
  library(package = "pheatmap")
  library(package = "FSA")
  library(package = "tidyverse")
})
```

```{r options}
options(stringsAsFactors       = FALSE,
	      width                  = 60,
        dplyr.summarise.inform = FALSE)
select <- dplyr::select
source(file = file.path(dataCacheDir, "hipc_fig_s4_helperFunctions.R"))
```

# Setup
These figures only look at the "Young" cohort that is defined as 18 to 50 years old and the "Extended Old" cohort of 50 to 90 years old.  The "Old" cohort (not used here) was 60 to 90 years old.
```{r read-esets}
young.NoResp <- readRDS(file = file.path(dataDir, "young_norm_eset.rds"))
young.WithResp <- readRDS(file = file.path(dataDir, "young_norm_withResponse_eset.rds"))
```

```{r set-standard-colors}
# Fig 1
set3colors <- brewer.pal("Set3", n = 12)
colors.fig1a <- c(set3colors[c(2, 11, 3, 12, 8)], 
                  "black", 
                  set3colors[c(7, 6, 5, 1, 4, 9, 10)])
# Other Figures
getPalette <- colorRampPalette(c(brewer.pal(name = "Set3", n = 10),
				                         brewer.pal(name = "Dark2", n = 4)))
vaccines <- unique(paste0(young.NoResp$pathogen, 
                          " (", 
                          young.NoResp$vaccine_type, 
                          ")"))
vaccine2color <- getPalette(n = length(vaccines)) %>%
  setNames(nm = sort(vaccines))
```

# Figures
## Figure 1. Creation of a combined dataset of transcriptional responses to vaccination across diverse vaccine platforms and target pathogens
### Figure 1b. Histogram of the timepoints pre- (days -7 and 0) and post-vaccination (days > 0) available in the immune signature data resource.
```{r fig1b, eval=TRUE}
plotDF <- pData(young.NoResp) %>%
  select(uid, study_time_collected, pathogen, vaccine_type) %>%
  arrange(study_time_collected) %>%
  mutate(study_time_collected = signif(study_time_collected, digits = 3),
	 # any sampling data after 20 days coded as >20
	 study_time_collected.factor = ifelse(test = study_time_collected > 20,
					                              yes  = ">20",
					                              no   = study_time_collected),
	 study_time_collected.factor =
	   factor(study_time_collected.factor,
		  levels = unique(study_time_collected.factor)),
	 vaccine = paste0(pathogen, " (", vaccine_type, ")")) %>%
  group_by(study_time_collected.factor, vaccine) %>%
  summarize(n = n())
# assign each vaccine a color
getPalette <- colorRampPalette(c(brewer.pal(name = "Set3", n = 10),
				 brewer.pal(name = "Dark2", n = 4)))
vaccine2color <- getPalette(n = length(unique(plotDF$vaccine))) %>%
  setNames(nm = sort(unique(plotDF$vaccine)))
ggplot(data = plotDF,
       mapping = aes(x = study_time_collected.factor,
		     y = n)) +
  geom_bar(stat = "identity", mapping = aes(fill = vaccine)) +
  labs(x = "Days post-vaccination", y = "Number of samples") +
  scale_y_continuous(limits = c(0, 800),
		     breaks = seq(from = 0, to = 800, by = 100)) + 
  scale_fill_manual(name   = "Pathogen (Vaccine type)",
		    values = vaccine2color) +
  theme_bw() +
  theme(panel.grid         = element_blank(),
	panel.grid.major.y = element_line(color = "lightgrey"),
	axis.text          = element_text(color = "black"),
	axis.text.x        = element_text(angle = 45, hjust = 1),
	legend.pos         = "bottom",
	legend.text        = element_text(size = 7),
	legend.key.height  = unit(0.02, units = "npc"),
	legend.key.width   = unit(0.015, units = "npc")) +
  guides(fill = guide_legend(ncol = 2))
```

### Figure 1c. Principal variance component analysis used to estimate the proportion of the variance observed in the transcriptomic data that can be attributed to clinical and experimental variables.
```{r boot-pvca, eval=FALSE}
mat <- exprs(young.NoResp)
pdata <- pData(young.NoResp) %>%
  mutate(Vaccine = paste0(pathogen, " (", vaccine_type, ")")) %>%
  select(y_chrom_present, ethnicity, Vaccine, study_time_collected,
	 age_imputed) %>%
  rename(Gender    = y_chrom_present,
	 Ethnicity = ethnicity,
	 Timepoint = study_time_collected,
	 Age       = age_imputed)
pdata <- pdata[colnames(mat), ]

B <- 1000
bootLS <- mclapply(1:B, FUN = function(seed) {
  set.seed(seed = seed)
  indices <- sample.int(n = ncol(mat), size = ncol(mat), replace = TRUE)
                 pdataTemp <- pdata[indices, ]
                 matTemp <- mat[, indices]
                 fit <- PVCA(counts    = matTemp,
                             meta      = pdataTemp,
                             threshold = 0.6,
                             inter     = FALSE)
  return(value = c(seed = seed, fit))
}, mc.cores = detectCores() - 1)
# save(bootLS, file = file.path(dataCacheDir, "fig1c.boot.rda"))
```

```{r fig1c, fig.width=4, fig.height=7, eval=TRUE}
load(file = file.path(dataCacheDir, "fig1c.boot.rda"))
plotDF <- bootLS %>%
  select(-seed) %>%
  apply(MARGIN = 2, FUN = quantile, probs = c(0.025, 0.5, 0.975)) %>%
  t() %>%
  as.data.frame() %>%
  mutate(effect = names(bootLS)[-1]) %>%
  arrange(`50%`) %>%
  mutate(effect = factor(effect, levels = effect))

ggplot(data    = plotDF,
       mapping = aes(x = effect, y = `50%`)) +
  geom_bar(stat = "identity") +
  geom_errorbar(mapping = aes(ymin = `2.5%`, ymax = `97.5%`), width = 0.1) +
  geom_text(aes(label = signif(`50%`, digits = 3)),
            nudge_y   = 0.03,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1))
```
          
Perform Sample Level Enrichment Analysis on hallmark+BTM genesets
Pre-Processing:
Microarray probes may have different affinities for their target mRNA and
thus the itensities can't be compared between probes inside the same sample.
A way to correct for that is to scale each probe to a mean of 0 and a standard-
deviation of 1.
The following looks only at pre-vaccination timepoints
```{r, check-cached-slea, eval=TRUE}
sleaFile <- file.path(dataCacheDir, "sleaSet.rda")
sleaCached <- file.exists(sleaFile)
if(sleaCached){
  load(file = sleaFile)
}
```

```{r, slea, eval=FALSE}
data("all_genesets", package = "ImmuneSignatures2")
batch <- interaction(young.NoResp$study_accession,
                     young.NoResp$matrix,
                     drop = TRUE)
scaledMat <- by(data = t(exprs(young.NoResp)),
                INDICES = batch,
                FUN = function(x) scale(x)) %>%
  do.call(what = rbind) %>%
  t()
scaledMat <- scaledMat[, sampleNames(young.NoResp)]
# SLEA (Ref: Lopez-Bigas N. et al. (2008) Genome Biol.)
B=100
flag <- lapply(all_genesets, FUN = intersect, y = rownames(scaledMat)) %>%
  sapply(FUN = length)
zscoreMat <- sapply(all_genesets, FUN = function(gs) {
  gs <- intersect(gs, rownames(scaledMat))
  ngenes <- length(gs)
  mu <- colMeans(scaledMat[gs, , drop = FALSE], na.rm = TRUE)
  muPermut <- mclapply(1:B, FUN = function(seed) {
    set.seed(seed = seed)
    muHat <- colMeans(scaledMat[sample.int(nrow(scaledMat), ngenes), , 
                                drop = FALSE],
                      na.rm = TRUE)
    return(value = muHat)
  })
  muPermut <- do.call(what = rbind, args = muPermut)
  zscore <- (mu - colMeans(muPermut, na.rm = TRUE)) /
    apply(muPermut, MARGIN = 2, FUN = sd, na.rm = TRUE)
  return(value = zscore)
})
```

```{r, create-slea-eset, eval=FALSE}
sleaSet <- ExpressionSet(assayData = t(zscoreMat),
                         phenoData = AnnotatedDataFrame(pData(young.NoResp)))
save(sleaSet, file = sleaFile)
```

```{r define-genesets, eval=TRUE}
inflamGS <- c("HALLMARK_INFLAMMATORY_RESPONSE",
              "HALLMARK_COMPLEMENT",
              "HALLMARK_IL6_JAK_STAT3_SIGNALING",
              "HALLMARK_TNFA_SIGNALING_VIA_NFKB")
```

## Figure 2. Heatmap of SLEA zscore on pre-vaccination samples
```{r, fig2, fig.width=8, fig.height=6, message=FALSE, eval=TRUE}
# filter SLEA zscore on pre-vaccination samples
zscoreTemp <- exprs(sleaSet)[, sleaSet$study_time_collected <= 0]
# remove genesets with missing symbols in virtual study
zscoreTemp <- zscoreTemp[rowMeans(is.na(zscoreTemp)) < 1, ]
# cutree into three groups (see gap statistic analysis
set.seed(seed = 17)
tree_col  <- kmeans(x = t(zscoreTemp), centers = 3)
gr <- tree_col$cluster
# relabel cluster based on inflammatory genesets
inflamLS <- c("HALLMARK_INFLAMMATORY_RESPONSE",
              "HALLMARK_COMPLEMENT",
              "HALLMARK_IL6_JAK_STAT3_SIGNALING",
              "HALLMARK_TNFA_SIGNALING_VIA_NFKB")
inflamDF <- data.frame(mu = apply(zscoreTemp[inflamLS, ], MARGIN = 2, FUN = median)) %>%
  rownames_to_column()
inflamDF <- data.frame(grn = gr) %>%
  rownames_to_column() %>%
  merge(x = inflamDF) %>%
  merge(y    = select(pData(sleaSet),
                      setdiff(varLabels(sleaSet), "gr")),
        by.x = "rowname",
        by.y = "uid")
inflamDF %>%
  group_by(grn) %>%
  summarize(q2 = median(mu)) %>%
  ungroup() %>%
  mutate(gr = ifelse(q2 %in% min(q2),
                     yes = "low",
                     no  = NA),
         gr = ifelse(q2 %in% max(q2) & is.na(gr),
                     yes = "high",
                     no  = gr),
         gr = ifelse(is.na(gr),
                     yes = "mid",
                     no  = gr),
         gr = factor(gr, levels = c("low", "mid", "high"))) %>%
  merge(x = inflamDF,
        by = "grn") %>%
  column_to_rownames(var = "rowname") -> inflamDF
inflamDF <- inflamDF[colnames(zscoreTemp), ]
rowLabels <- rownames(zscoreTemp)
rowLabels[!grepl(pattern = "B.cell|E2F|T.cell|MYC| NK",
		 rownames(zscoreTemp),
		 ignore.case = TRUE) &
	    !grepl(pattern = "Interferon|STAT|migration|Monocytes| DC|Neutrophiles",
		   rownames(zscoreTemp),
                 ignore.case = TRUE) &
            !(rownames(zscoreTemp) %in% inflamLS)] <- ""
# generate heatmap annotation
ha <- inflamDF %>%
  rownames_to_column() %>%
  mutate(inflam.gs = findInterval(mu,
				  vec = quantile(mu ,
						 probs = c(0, 0.33, 0.66, 1)),
				  rightmost.closed= TRUE),
	 inflam.gs = paste0("tier", inflam.gs)) %>%
  select(rowname, inflam.gs) %>%
  column_to_rownames() %>%
  .[colnames(zscoreTemp), , drop=FALSE] %>%
  HeatmapAnnotation(df = ., col = list(inflam.gs = c(tier1 = "yellow",
tier2 = "orange",
 tier3 = "red")))
heat <- Heatmap(matrix                      = zscoreTemp[!(rowLabels %in% ""), 
                                                         ],
	row_names_side              = "left", 
	show_column_names           = FALSE,
	column_split                = inflamDF$gr,
	show_row_dend               = FALSE,
	name                        = "SLEA z-score",
	#row_labels                  = rowLabels,
	row_names_gp                = gpar(fontsize = 8),
	top_annotation              = ha)
ht <- draw(heat)
heatColOrder <- column_order(ht)
```

```{r, append-gr-sleaset, eval=FALSE}
sleaSet$gr <- inflamDF[match(sleaSet$uid,
                             table = rownames(inflamDF)), "gr"]
```

```{r, generate-pts-gr, eval=TRUE}
inflamGr <- pData(sleaSet) %>%
  arrange(desc(study_time_collected)) %>%
  filter(!is.na(gr)) %>%
  select(participant_id,
         gr) %>%
  distinct() %>%
  filter(!duplicated(participant_id)) %>%
  mutate(gr = as.vector(gr)) %>%
  rename(inflam.gr = gr)
```

Create pathway specific gene-level heatmaps
```{r, read-generif, eval=TRUE}
# use generif to select top 10
generifPath <- file.path(dataCacheDir, "generifs_basic")
generif <- scan(file = generifPath, what = "raw", sep = "\n")
generif <- generif %>%
  strsplit(split = "\t") %>%
  do.call(what = rbind)
header <- generif[1, ] %>%
  gsub(pattern = "#", replacement = "")
generif <- generif[-1, ] %>%
  as.data.frame() %>%
  setNames(header)
human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
batch <- interaction(young.NoResp$study_accession,
                     young.NoResp$matrix,
                     drop = TRUE)                                                          
scaledMat <- by(data = t(exprs(young.NoResp)),
                INDICES = batch,
                FUN = function(x) scale(x)) %>%
  do.call(what = rbind) %>%
  t()                                                                                      
scaledMat <- scaledMat[, sampleNames(young.NoResp)] 
```

```{r, fig2.1, eval=TRUE, message=FALSE}
nkGenes <- all_genesets[grepl(pattern = " NK", names(all_genesets))]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(nkGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = " NK|Natural Killer", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
col_fun = colorRamp2(seq(-3, 3, length.out = 5),
                     heat@matrix_color_mapping@colors)
heatNK <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "NK")
draw(heatNK)
```

```{r, fig2.2, eval=TRUE}
tGenes <- all_genesets[grepl(pattern = "T.cell", names(all_genesets))]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(tGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "T.cell|Lymphocytes T", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatT <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "T")
draw(heatT)
```

```{r, fig2.3, eval=TRUE}
meGenes <- all_genesets[grepl(pattern = "E2F|MYC", names(all_genesets))]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(meGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "MYC|E2F", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatE2F <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "MYC|E2F")
draw(heatE2F)
```

```{r, fig2.4, eval=TRUE}
bGenes <- all_genesets[grepl(pattern = "B.cell", names(all_genesets)) &
              !grepl(pattern = "TBA", names(all_genesets))]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(bGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "B.cell|Lymphocytes B", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatB <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "B")
draw(heatB)
```

```{r, fig2.5, eval=TRUE}
mdGenes <- all_genesets[grepl(pattern = "Monocytes|DC", names(all_genesets), ignore.case = TRUE)]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(mdGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "Monocytes|Dendritic", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatMono <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "Monocytes|DC")
draw(heatMono)
```

```{r, fig2.6, eval=TRUE}
iGenes <- all_genesets[inflamLS]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(iGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "Inflammat", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatInflam <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "Inflammation")
draw(heatInflam)
```

```{r, fig2.7, eval=TRUE}
isgGenes <- all_genesets[grepl(pattern = "Interferon|STAT", names(all_genesets), ignore.case = TRUE)]
symbol2id <- getBM(attributes  = c("hgnc_symbol", "entrezgene_id"),
                   filters     = "hgnc_symbol",
                   values = unique(unlist(isgGenes)),
                   mart        = human)
generifTemp <- merge(x = symbol2id,
      y     = generif,
      by.x  = "entrezgene_id",
      by.y = "Gene ID",
      all.x = TRUE) %>%
  filter(grepl(pattern = "Interferon", `GeneRIF text`, ignore.case = TRUE)) %>%
    filter(hgnc_symbol %in% featureNames(young.NoResp)) %>%
    group_by(hgnc_symbol) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    top_n(n = 10, wt = n)
# draw heatmap column same order as figure 2
mat <- scaledMat[generifTemp$hgnc_symbol, colnames(zscoreTemp)[unlist(heatColOrder)]]
heatISG <- Heatmap(matrix = mat, name = "z-score", 
        col = col_fun,
        cluster_columns = FALSE,
        column_split  = factor(rep(names(heatColOrder), sapply(heatColOrder, FUN = length)), levels = names(heatColOrder)),
        show_row_dend = FALSE,
        show_column_names = FALSE,
        row_title = "Interferons")
draw(heatISG)
```

## Figure 3. Kinetics of the vaccine response are dictated by the pre-vaccination endotypes
### Figure 3a. Line plots showing the expression of inflammatory pathways
Plot SLEA z-score of inflammatory genesets over time
```{r fig3a, fig.width=6.92, fig.height=2.96, warning=FALSE, eval=TRUE}
sleaSet$gr <- inflamGr$inflam.gr[match(sleaSet$participant_id,
                                       table = inflamGr$participant_id)]

plotDF <- data.frame(mu = colMeans(exprs(sleaSet)[inflamGS, ])) %>%
  rownames_to_column(var = "uid") %>%
  merge(y = pData(sleaSet), by = "uid") %>%
  mutate(gr = recode(gr,
                     "low"  = "inflam.lo",
                     "mid"  = "inflam.mid",
                     "high" = "inflam.hi"),
         gr = factor(gr,
                     levels = c("inflam.lo",
                                "inflam.mid",
                                "inflam.hi"))) %>%
  filter(!is.na(gr))
ggplot(data = plotDF,
       mapping = aes(x = study_time_collected, y = mu, color = gr)) +
  geom_line(mapping = aes(group = participant_id),
            alpha = 0.1) +
  geom_hline(yintercept = 0, color = "grey", size = 2) +
  geom_smooth(method = "loess", formula = "y~x", color = "black") +
  facet_wrap(facets = ~gr) +
  scale_x_continuous(limits = c(0, 28)) +
  labs(y = "Inflammatory pathways\n(SLEA z-score)",
       x = "Time after vaccination (days)") +
  theme_bw() +
  theme(axis.text    = element_text(size = 15),
        strip.text   = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.pos   = "none",
        panel.grid  = element_blank())
```

### Figure 3b. Line plots showing the expression of ISGs
Plot SLEA z-score of interferon genesets over time
```{r, fig3b, fig.width=6.92, fig.height=2.96, warning=FALSE, eval=TRUE}
isgLS  <- c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
            "HALLMARK_INTERFERON_GAMMA_RESPONSE")
plotDF <- data.frame(mu = colMeans(exprs(sleaSet)[isgLS, ])) %>%
  rownames_to_column(var = "uid") %>%
  merge(y = pData(sleaSet), by = "uid") %>%
  mutate(gr = recode(gr,
                     "low"  = "inflam.lo",
                     "mid"  = "inflam.mid",
                     "high" = "inflam.hi"),
         gr = factor(gr,
                     levels = c("inflam.lo",
                                "inflam.mid",
                                "inflam.hi"))) %>%
  filter(!is.na(gr))
ggplot(data = plotDF,
       mapping = aes(x = study_time_collected, y = mu, color = gr)) +
  geom_line(mapping = aes(group = participant_id),
            alpha = 0.1) +
  geom_hline(yintercept = 0, color = "grey", size = 2) +
  geom_smooth(method = "loess", formula = "y~x", color = "black") +
  facet_wrap(facets = ~gr) +
  scale_x_continuous(limits = c(0, 28)) +
  labs(y = "Interferon-stimulated genes\n(SLEA z-score)",
       x = "Time after vaccination (days)") +
  theme_bw() +
  theme(axis.text    = element_text(size = 15),
        strip.text   = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.pos   = "none",
        panel.grid  = element_blank())
```

### Figure 3c. Line plots showing the expression of B cells markers
Plot SLEA z-score of B cell genesets over time
```{r, fig3c, fig.width=6.92, fig.height=2.96, warning=FALSE, eval=TRUE}
bLS <- c("enriched in B cells (I) (M47.0)",
         "enriched in B cells (V) (M47.4)",
         "plasma cells & B cells, immunoglobulins (M156.0)")
plotDF <- data.frame(mu = colMeans(exprs(sleaSet)[bLS, ])) %>%
  rownames_to_column(var = "uid") %>%
  merge(y = pData(sleaSet), by = "uid") %>%
  mutate(gr = recode(gr,
                     "low"  = "inflam.lo",
                     "mid"  = "inflam.mid",
                     "high" = "inflam.hi"),
         gr = factor(gr,
                     levels = c("inflam.lo",
                                "inflam.mid",
                                "inflam.hi"))) %>%
  filter(!is.na(gr))
ggplot(data = plotDF,
       mapping = aes(x = study_time_collected, y = mu, color = gr)) +
  geom_line(mapping = aes(group = participant_id),
            alpha = 0.1) +
  geom_hline(yintercept = 0, color = "grey", size = 2) +
  geom_smooth(method = "loess", formula = "y~x", color = "black") +
  facet_wrap(facets = ~gr) +
  scale_x_continuous(limits = c(0, 28)) +
  labs(y = "B cells\n(SLEA z-score)",
       x = "Time after vaccination (days)") +
  theme_bw() +
  theme(axis.text    = element_text(size = 15),
        strip.text   = element_text(size = 20),
        axis.title.x = element_text(size = 20),
        legend.pos   = "none",
        panel.grid  = element_blank())
```

## Figure 4. Prediction of the antibody response by the pre-vaccination endotypes
### Figure 4a. Boxplot of the MFC as a function of the pre-vaccination inflammation endotypes. 
```{r, fig4a, warning=FALSE, fig.width=4.4, fig.height=4.6, eval=TRUE}
load(file = sleaFile)
inflamDF <- pData(sleaSet)[sleaSet$study_time_collected <= 0, ]
plotDF <- inflamDF %>%
  merge(y = distinct(select(pData(young.WithResp), 
                            participant_id, 
                            MFC,
                            MFC_p30))) %>%
  group_by(study_accession) %>%
  mutate(sMFC = scale(MFC),
         vaccine = paste0(pathogen,".", vaccine_type),
         vaccine = ifelse(pathogen %in% "Meningococcus",
                          yes = "Meningococcus",
                          no  = vaccine))
ggplot(data = plotDF,
       mapping = aes(x = factor(gr), y = sMFC)) +
       geom_beeswarm(cex = 1.1, size = 0.75) +
       geom_boxplot(outlier.color = "transparent", 
                    fill = "transparent", 
                    col = "red") +
       labs(x = "Prevax inflammatory cluster") +
       theme_bw()
kruskal.test(x = plotDF$sMFC, g = plotDF$gr)
pairwise.wilcox.test(x = plotDF$sMFC, g = plotDF$gr, p.adjust.method = "none") %>%
    print()
```

### Figure 4b. ROC curve estimated from the 10-fold CV
Load predictions
```{r, load-prediction, eval=TRUE}
predFile<- file.path(dataCacheDir, 
                     "random_forest_output/DB.PREDS.rda")
load(file = predFile)
```

```{r, fig4b, eval=TRUE}
par(pty = "s")

### 10-CV 
ROC.OBJ1 <- roc(response      = db.preds$obs, 
                predictor     = db.preds$highResponder, 
                levels        = c("lowResponder", "highResponder"),
                direction     = "<",
                plot          = TRUE, 
                legacy.axes   = TRUE, 
                percent       = TRUE,
                xlab          = "False Positive %",
                ylab          = "True Positive %",
                identity      = FALSE,
                col           = "black",
                lwd           = 4, 
                lty           = 1, 
                print.auc.x   = 77,
                print.auc.y   = 28,
                print.auc.cex = 0.9,
                ci            = TRUE, 
                print.auc     = TRUE)

### Training
ROC.OBJ2 <- roc(response      = preds.train$obs, 
                levels        = c("lowResponder", "highResponder"),
                predictor     = preds.train$highResponder, 
                direction     = "<",
                legacy.axes   = TRUE, 
                percent       = TRUE,
                ci            = TRUE)

plot.roc(ROC.OBJ2,
         legacy.axes   = TRUE, 
         percent       = TRUE,
         col           = "gray60",
         lwd           = 4,
         lty           = 3,
         print.auc.x   = 85,
         print.auc.y   = 21,
         print.auc.cex = 0.9,
         ci            = TRUE,
         print.auc     = TRUE, 
         add           = TRUE)

legend("bottomright", 
       legend = c("Training", "10-CV"),
       col    = c("gray60", "black"),
       lty    = c(3, 1),
       lwd    = 4,
       cex    = 0.8,
       ncol   = 2)
```

### Figure 4c. Heatmap showing the pre-vaccination expression of the overlapping inflammatory genes
Read features and importance
```{r, read-feat-sup, warning=FALSE, message=FALSE, eval=TRUE}
featFile <- file.path(dataCacheDir,
                      "Top500_Var_VarImp.csv")
featDF <- read_csv(file = featFile, 
                   col_names = c("FEATURE", "IMPORTANCE"),
                   skip = 1) %>%
  filter(IMPORTANCE >= 0)
```

Fisher exact test on hallmark genesets
```{r, fisher, warning=FALSE, eval=TRUE}
data("all_genesets", package = "ImmuneSignatures2")

# fisher exact test to test overlap
bg <- featureNames(young.WithResp)
fisherDF <- mclapply(names(all_genesets), FUN = function(gsName) {
  gs <- all_genesets[[gsName]]
  tab <- table(factor(bg %in% gs, levels = c(TRUE, FALSE)),
               factor(bg %in% featDF$FEATURE, levels = c(TRUE, FALSE)))
  fit <- fisher.test(tab, alternative = "greater")
  return(value = data.frame(gsName = gsName,
         or    = unname(fit$estimate),
         p     = fit$p.value,
         inter = paste(intersect(gs, featDF$FEATURE), collapse = ",")))
}, mc.cores = detectCores() - 1) %>%
  do.call(what = rbind) %>%
  mutate(adj.p = p.adjust(p, method = "BH"))

# print only genesets in Figure 2
filter(fisherDF, 
       adj.p <= 0.05 &
        (grepl(pattern     = "B.cell|E2F|T.cell|MYC| NK",
               gsName,
               ignore.case = TRUE) |
         grepl(pattern     = "Interferon|STAT|Inflam|Compl|TNF|NfkB",
               gsName,
               ignore.case = TRUE) |
         grepl(pattern     = "migration|Monocytes| DC|Neutrophiles",
               gsName,
               ignore.case = TRUE))) %>%
  kable()
```

Heatmap of inflammatory genes
```{r, fig4c, fig.width=8, fig.height=8, eval=TRUE}
inflamSigDF<- filter(fisherDF, gsName %in% inflamGS & adj.p <= 0.05)

rowAnnot <- inflamSigDF$inter %>%
  strsplit(split = ",") %>%
  setNames(inflamSigDF$gsName) %>%
  stack() %>%
  mutate(flag = 1) %>%
  pivot_wider(names_from = ind, values_from = flag, values_fill = 0) 

# append feature importance
featDF %>% 
  group_by(FEATURE) %>% 
  summarize(IMPORTANCE = mean(IMPORTANCE)) %>%
  merge(x    = rowAnnot,
        by.x = "values",
        by.y = "FEATURE") %>%
  column_to_rownames(var = "values") %>%
  arrange(desc(HALLMARK_TNFA_SIGNALING_VIA_NFKB),
          desc(HALLMARK_INFLAMMATORY_RESPONSE), 
          desc(HALLMARK_IL6_JAK_STAT3_SIGNALING),
          desc(IMPORTANCE)) -> rowAnnot

# extract gene expression
mat <- exprs(young.WithResp)[rownames(rowAnnot),
             young.WithResp$study_time_collected <= 0 &
             young.WithResp$MFC_p30 %in% c("lowResponder",
"highResponder")]
mat <- t(scale(t(mat)))

colAnnot <- pData(young.WithResp)[, "MFC_p30", drop = FALSE]
rownames(colAnnot) <- young.WithResp$uid
colAnnot <- colAnnot[colnames(mat), , drop = FALSE]
colOrder <- order(colMeans(mat)) 
ha <- HeatmapAnnotation(df  = colAnnot[colOrder, , drop= FALSE],
                        col = list(MFC_p30 = c(lowResponder  = "black",
                                               highResponder = "red")),
                        annotation_legend_param = list(title_gp = gpar(fontsize = 8),
                                    labels_gp = gpar(fontsize = 8)))
ra <- rowAnnotation(df  = rowAnnot,
                    col = list(HALLMARK_INFLAMMATORY_RESPONSE =
                                 c("1" = "black", "0" = "white"),
                               HALLMARK_TNFA_SIGNALING_VIA_NFKB =
                                 c("1" = "black", "0" = "white"),
                               HALLMARK_IL6_JAK_STAT3_SIGNALING =
                                 c("1" = "black", "0" = "white")),
                    annotation_legend_param = list(title_gp = gpar(fontsize = 8),
                                    labels_gp = gpar(fontsize = 8)),
                    annotation_name_gp = gpar(fontsize = 8))
ht <- Heatmap(matrix            = mat[rownames(rowAnnot), colOrder],
        cluster_rows      = FALSE,
	      row_names_side    = "left", 
	      show_column_names = FALSE,
        cluster_columns   = FALSE,
	      name              = "z-score",
        row_names_gp      = gpar(fontsize = 6),
        heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                                    labels_gp = gpar(fontsize = 8)),
	      top_annotation    = ha,
        left_annotation   = ra)
print(ht)
wilcox.test(formula = colOrder~MFC_p30,
            data    = cbind(colAnnot, colOrder = colMeans(mat)))
```

### Figure 4d. Comparison of seven genes contributing the majority of the classifier prediction (importance > 50%) against previously identified pre-vaccination signatures of vaccine response
Load data and signatures
```{r, load-dashboard-sig, eval=TRUE}
# reproducible results for permutation testing
set.seed(seed = 021093)

# load data
# data.frame built from ImmPort linking SDY and PMIDs
train_test_df <- readRDS(file = file.path(dataCacheDir, "ImmPort_PMID2SDY.RDS"))  
# data frame of HIPC baseline signatures, manually annotated pairing of up/down genes
filtered_resp_sigs <- readRDS(file = file.path(dataCacheDir, 
                                               "fig4_response_signatures.RDS")) 
# read feature from RF model (Fig 4bc)
featFile <- file.path(dataCacheDir, "Top500_Var_VarImp.csv")
# use gene feature present in at least 50% of the models obtain by bagging
geneUp <- read_csv(file = featFile, 
                   col_names = c("FEATURE", "IMPORTANCE"),
                   skip = 1) %>%
  filter(IMPORTANCE >= 50) %>%
  .$FEATURE %>%
  paste(collapse = ";")
# update list of pre-vax signature with the current study
filtered_resp_sigs$"response.component.(gene.symbol)"[
  filtered_resp_sigs$"publication.reference.(PMID)" %in% "current study" &
    filtered_resp_sigs$"response.behavior" %in% "up-regulated for"] <- 
  geneUp
filtered_resp_sigs$paired[
  filtered_resp_sigs$"publication.reference.(PMID)" %in% "current study" &
    filtered_resp_sigs$"response.behavior" %in% "up-regulated for"] <- "F"
filtered_resp_sigs <- filtered_resp_sigs[!(
  filtered_resp_sigs$"publication.reference.(PMID)" %in% "current study" &
    filtered_resp_sigs$"response.behavior" %in% "down-regulated for"), ]
```

Prepare data for analysis
```{r, prep-data, results='hide', eval=TRUE}
# convert virtualStudy ExpressionSet to MetaIntegrator dataset objects
MIobj_list <- data2MIobj(data = young.WithResp)

# convert signatures to MetaIntegrator filter objects
filter_list <- signatures2Filters(filtered_resp_sigs) 

# identify data (SDY) and signature (PMID) pathogens
study_pathogens <- sapply(MIobj_list, FUN = function(x) {
  return(value = unique(x$pheno$pathogen))
  }) %>%
  unlist() %>%
  unique()
filter_pathogens <- sapply(filter_list, FUN = function(x) {
  strsplit(x$filterDescription, split = '_') %>%
    unlist() %>%
    dplyr::first()
}) %>%
  unlist() %>%
  unique()

# remove signatures that don't map onto measured genes or from elderly adults
filter_remove_ix <- sapply(filter_list, function(x) {
  # signatures where greater than 50% of genes are mapped in virtualStudy
  gene_match_ix <- c(x$posGeneNames, x$negGeneNames) %>%
    setdiff(y = "") %>%
    (function(x) mean(x %in% featureNames(young.WithResp)) < 0.5)
  # signatures that are empty
  length_ix <- length(setdiff(x$posGeneNames, "")) == 0 & 
    length(setdiff(x$negGeneNames, "")) == 0
  # signatures identified in seniors
  age_ix <- grepl(pattern = 'senior', x = x$filterDescription)
  return(value = gene_match_ix | length_ix | age_ix)
})
message("number of signature to be tested:")
sum(!filter_remove_ix)
filter_list <- filter_list[!filter_remove_ix]
```

Evaluate dashboard pre-vax signatures in virtual study
```{r, eval-dashboard-prevax-sig, eval=TRUE}
# evaluate signatures in data
subject_plot_dat_list <- list()

for (j in 1:length(MIobj_list)) {
  current_study <- MIobj_list[[j]]
  for (i in 1:length(filter_list)) {
    message(paste0("study=", 
                   j, 
                   "/", 
                   length(MIobj_list), 
                   " - sig=", 
                   i, 
                   "/", 
                   length(filter_list)))
    
    current_filter <- filter_list[[i]]
    subject_plot_dat_obj <- data.frame('Study'           = current_study$formattedName,
                                       'Pathogen'        = current_study$pheno$pathogen,
                                       'Filter'          = current_filter$filterDescription,
                                       'Filter_index'    = i,
                                       'Filter_Pathogen' = strsplit(current_filter$filterDescription, split = '_')[[1]][1],
                                       'SubjectID' = current_study$pheno$participant_id,
                                       'Response'  = as.factor(current_study$class),
                                       'Score'     = calculateScoreRobust(filterObject = current_filter, datasetObject = current_study, suppressMessages = TRUE),
                                       stringsAsFactors = FALSE)
    
    # check number of genes from signature
    filter_genes <- setdiff(c(current_filter$posGeneNames, current_filter$negGeneNames), '')
    fraction_matched <- mean(filter_genes %in% current_study$keys)
    if(fraction_matched >= 0.5){
      auroc <- calculateROC(labels = as.numeric(subject_plot_dat_obj$Response), 
                            predictions = subject_plot_dat_obj$Score, 
                            AUConly = T)[1]
      pval <- calculatePval(labels = as.numeric(subject_plot_dat_obj$Response), 
                            values = subject_plot_dat_obj$Score, 
                            true_value = auroc, n_perm = 1000)
    } else {
      auroc <- NA
      pval <- NA
    }
    
    
    subject_plot_dat_obj <- subject_plot_dat_obj %>%
      mutate('AUROC' = auroc, 'P.Value' = pval) 
    
    subject_plot_dat_list[[paste0(current_study$formattedName, '_', current_filter$filterDescription, '_', i)]] <- subject_plot_dat_obj
  }
}
```

Plot formatting
```{r, plot-format, eval=TRUE}
# get study sizes for plot ordering
study_sizes <- sapply(MIobj_list, function(x) {
  return(x$N)
})
study_sizes <- sort(study_sizes, decreasing = TRUE)

# make subject-level data frame of all signature scores
subject_plot_dat <- bind_rows(subject_plot_dat_list) %>%
  mutate(Study = factor(Study, levels = names(study_sizes), ordered = TRUE)) %>%
  mutate('Filter.Behavior' = sapply(Filter, function(x){str_split(string = x, pattern = '_')[[1]][2]})) %>%
  mutate('Filter.Behavior' = gsub(pattern = 'positively', replacement = 'up', x = Filter.Behavior)) %>%
  mutate('Filter.Behavior' = gsub(pattern = 'negatively', replacement = 'down', x = Filter.Behavior)) %>%
  mutate(Filter.Behavior = ifelse(grepl(pattern = 'up', x = Filter.Behavior), 'up', Filter.Behavior)) %>%
  mutate(Filter.Behavior = ifelse(grepl(pattern = 'down', x = Filter.Behavior), 'down', Filter.Behavior)) %>%
  mutate(Filter = paste0(Filter, '_', Filter_index))

# standardize influenza virus --> influenza
influenza_ix <- grepl(pattern = 'influenza', x = subject_plot_dat$Filter_Pathogen)
subject_plot_dat$Filter_Pathogen[influenza_ix] <- 'influenza'

# annotate sdy and pmid
study_plot_dat <- subject_plot_dat %>%
  dplyr::select(Study, Pathogen, Filter, Filter_index, AUROC, Filter_Pathogen, P.Value, Filter.Behavior) %>%
  distinct() %>%
  mutate('Train' = F) %>%
  mutate('SDY' = sapply(X = Study, FUN = function(x){str_split(string = x, pattern = '_')[[1]][1]})) %>%
  mutate('PMID' = sapply(X = Filter, FUN = function(x){str_split(string = x, pattern = '_')[[1]][3]}))
```

Label signature discovery datasets using ImmPort Data Frame
```{r, label-sig, results='hide', eval=TRUE}
# annotate which datasets were used as discovery for each signature
train_test_vect <- study_plot_dat %>%
  dplyr::select(SDY, PMID, Train)
for (i in 1:nrow(train_test_vect)){
  current_eval <- train_test_vect[i,]
  relevant_publications <- train_test_df %>% 
    filter(SDY == current_eval$SDY)
  # if we find a signature PMID that matches a dataset PMID, mark $Train = T
  if(nrow(relevant_publications) > 0){
    if(current_eval$PMID %in% relevant_publications$Publication){
      current_eval$Train <- T
      print(paste(current_eval$PMID, relevant_publications$Publication))
    } else {
      current_eval$Train <- F
    }
  }
  train_test_vect[i,] <- current_eval
}
study_plot_dat$Train <- train_test_vect$Train

# create symbols for train/test & p.value significance plot overlay
study_plot_dat <- study_plot_dat %>%
  mutate(Train_Label = ifelse(Train, '○', '')) %>%
  mutate(Train_Label = ifelse(PMID == 'current study', '○', Train_Label)) %>%
  mutate(Significance_Label = ifelse(P.Value <= 0.05, '*', ''))

study_plot_dat <- study_plot_dat %>%
  mutate('Vaccine.Type' = sapply(X = Study, function(x){strsplit(x = as.character(x), split = '_')[[1]][3]}))
```

Plot label clean-up
```{r, label-clean-up, results='hide', eval=TRUE}
# format study names for plot axes
new_study_names <- as.character(unique(study_plot_dat$Study)) %>%
  gsub(pattern = paste(paste0('_', c('Influenza', 'Yellow Fever', 'Pneumococcus', 'Smallpox', 'Tuberculosis', 'Hepatitis B', 'Meningococcus', 'Varicella Zoster'), '_'), collapse = '|'), replacement = '') %>%
  gsub(pattern = paste(c('Inactivated', 'Live attenuated', 'Conjugate', 'Polysaccharide', 'Recombinant Viral Vector'), collapse = '|'), replacement = '') %>%
  gsub(pattern = '_', replacement = ' ')
name_swap_df <- data.frame('Study' = unique(as.character(study_plot_dat$Study)), 
                           'New_Study' = new_study_names,
                           'Factor_Study' = unique(study_plot_dat$Study)) %>%
  arrange(Factor_Study) %>%
  mutate(Study = Factor_Study) %>%
  dplyr::select(-Factor_Study)
new_study_levels <- name_swap_df %>%
  dplyr::select(New_Study) %>%
  distinct()
name_swap_df <- name_swap_df %>%
  mutate(New_Study = factor(New_Study, levels = new_study_levels$New_Study, ordered = T))

# get sample sizes for plot
sample_N <- subject_plot_dat %>%
  dplyr::select(Study, SubjectID) %>%
  distinct() %>%
  group_by(Study) %>%
  summarize(N = n()) %>%
  ungroup() %>%
  mutate(Study = as.character(Study)) 

# add sample sizes & formatted plot names, reformat PMID 32094927 pathogen (previously overwritten)
study_plot_dat <- study_plot_dat %>%
  left_join(sample_N, by = 'Study') %>%
  left_join(name_swap_df, by = 'Study') %>%
  mutate(Study = New_Study) %>%
  mutate(Filter_Pathogen = ifelse(PMID == '32094927', 'influenza & yellow fever', Filter_Pathogen)) 

# re-order pathogens in plot
extra_paths <- c('Influenza', 'Yellow Fever')
path_order <- unique(study_plot_dat$Pathogen) %>%
  sort() %>%
  setdiff(extra_paths) 
path_order <- c(extra_paths, path_order)

study_plot_dat <- study_plot_dat %>%
  filter(!is.na(AUROC)) %>%
  mutate(PMID = factor(PMID, levels = sort(unique(PMID), decreasing = T), ordered = T),
         Filter.Behavior = factor(Filter.Behavior, levels = c('up', 'down'), ordered = T),
         Filter_Pathogen = factor(Filter_Pathogen, levels = c('hepatitis B virus', 'influenza', 'influenza & yellow fever', 'all vaccines'), ordered = T),
         Vaccine.Type = gsub(pattern = 'Rec', replacement = '        Rec', x = Vaccine.Type),
         Pathogen = factor(Pathogen, levels = path_order, ordered = T),
         N_Lab = ifelse(Filter == "hepatitis B virus_up-regulated for_30205979_9", as.character(N), ''))
# sample size generate label for top row of plot
label_dat <- study_plot_dat %>%
  mutate('Lab' = ifelse(Filter == "hepatitis B virus_up-regulated for_30205979_3", as.character(N), ''))

table(label_dat$Lab == '', label_dat$Filter)
```

Generate plots
```{r, fig4d, fig.height=8, fig.width=10, eval=TRUE}
p0 <- ggplot(study_plot_dat, 
             aes(x = Study, y= Filter, fill = AUROC, label = Significance_Label)) + 
  geom_tile() +
  geom_text(aes(label = Significance_Label), position = position_nudge(y = -0.27), size = 8) +
  geom_text(aes(label = Train_Label), position = position_nudge(y = 0.08), size = 8) + 
  geom_text(aes(label = N_Lab)) + 
  scale_fill_gradient2(midpoint = 0.5, low = scales::muted('blue'), high = scales::muted('red')) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(strip.text.y.right = element_text(angle = 0, margin = margin(t = 0, r = 0, b = 0, l = 0) )) +
  theme(strip.text.x.top = element_text(angle = 90, margin = margin(t = 0, r = 0, b = 0, l = 0))) + 
  theme(axis.text.y = element_blank()) + 
  theme(panel.spacing.y = unit(0, 'lines')) +
  theme(legend.position = 'left') + 
  facet_grid(Filter_Pathogen + PMID ~ Pathogen + Vaccine.Type, 
             scales = 'free', 
             space= 'free') +
  labs(y = 'Signature')
print(p0)
```

## Figure 5. Pre-vaccination endotypes in scRNAseq
### Figure 5a. UMAP of PBMCs from 20 healthy participants profiled by CITE-seq
Load CITE-seq data from SDY80 (Kotliarov et. al.) and normalize and denoise protein data with dsb. add adjmfc response data. 
```{r, load-citeseq, eval=TRUE}
set.seed(seed = 1234)

# read raw data and dsb norm by batch. 
s <- readRDS(file = file.path(dataCacheDir, 
                              "CITEseq_PMID32094927_seurat2.4hipcII_processed.rds"))
s <- UpdateSeuratObject(s)
```

Run umap based on normalized protein expression (load pre-computed umap embeddings ; code also provided below) umap was run by calling python via reticulate
```{r, load-umap, eval=TRUE}
# library(reticulate); use_virtualenv("r-reticulate")
# library(umap)
# 
# # set umap config
# config = umap.defaults
# config$n_neighbors = 40
# config$min_dist = 0.4
# 
# # run umap
# ump = umap(t(s@assay$CITE@data), config = config)
# umap_res = as.data.frame(ump$layout)
# colnames(umap_res) = c("UMAP_1", "UMAP_2")

umap_res <- readRDS(file = file.path(dataCacheDir,
                                     "umap_embeddings.rds"))
```

umap plot of protein based cell types
```{r, fig5a, fig.height=8, fig.width=8, eval=TRUE}
# combined data frame
d <- cbind(s@meta.data, t(as.data.frame(s$CITE@data)), umap_res)

# umap labeled. 
centers <- d %>% 
  group_by(celltype_label_3) %>% 
  summarize(UMAP_1 = median(UMAP_1), 
            UMAP_2 = median(UMAP_2))
centers$celltype_label_3 <- str_replace_all(string      = centers$celltype_label_3, 
                                            pattern     = "_",
                                            replacement = " ")
centers <- centers %>% 
  filter(celltype_label_3 != "DOUBLET")
d$celltype_label_3 <- str_replace_all(string      = d$celltype_label_3, 
                                      pattern     = "_",
                                      replacement = " ")

# color vec
cu <- c(
  "#89C5DA", "#DA5724", "#74D944", "#CE50CA", "#3F4921", "#C0717C", "#CBD588", "#5F7FC7",
  "#673770", "#D3D93E", "#38333E", "#508578", "#D7C1B1", "#689030", "#AD6F3B", "#CD9BCD",
  "#D14285", "#6DDE88", "#652926", "#7FDCC0", "#C84248", "#8569D5", "#5E738F", "#D1A33D",
  "#8A7C64", "#599861")

plot_theme = list(theme_bw(), 
                  theme(panel.grid.minor = element_blank(),
                        panel.grid.major = element_blank(),
                        axis.ticks.y     = element_blank(), 
                        axis.ticks.x     = element_blank())) 

# plot umap with cluster labels 
p <- ggplot(d, aes(x = UMAP_1,y =UMAP_2, color = celltype_label_3)) +
  plot_theme + 
  geom_point(show.legend = FALSE, size = 0.6, shape = 16,  alpha = 0.8) +
  scale_color_manual(values = cu) +
  ggrepel::geom_text_repel(data = centers, seed = 1, color = "black", box.padding = 1,
                  fontface = "bold", size = 4,
                  segment.color = "grey", 
                  segment.size = 0.3,force = 8,
                  aes(label = celltype_label_3),
                  show.legend = FALSE) 
print(p)
```

### Figure 5b. Single cell CITE-seq deconvolution of inflammatory genes
Single cell mapping of the ensemble classifier signature.  
```{r, fig5b, fig.width=8, fig.height=3, eval=TRUE}
# read feature from RF model (Fig 4bc)
featFile <- file.path(dataCacheDir,
                      "Top500_Var_VarImp.csv")
featDF <- read_csv(file = featFile, 
                   col_names = c("FEATURE", "IMPORTANCE"),
                   skip = 1) %>%
  filter(IMPORTANCE >= 0)

# read inflammatory genes (Fig 2)
# read hallmark
inflamLS <- c("HALLMARK_INFLAMMATORY_RESPONSE",
              # "HALLMARK_COMPLEMENT", complement was not significant in Fig 4
              "HALLMARK_IL6_JAK_STAT3_SIGNALING",
              "HALLMARK_TNFA_SIGNALING_VIA_NFKB")
inflamSig<- intersect(featDF$FEATURE, unlist(all_genesets[inflamLS]))

sigsub <- intersect(inflamSig, rownames(s$RNA@counts))

# normalize rna data with library size scaling factors 
s <- s %>% 
  NormalizeData(normalization.method = "LogNormalize") %>% 
  ScaleData(features = sigsub)

# map to high resolution protein phenotypes 
d3 <- cbind(s@meta.data, as.data.frame(t(as.matrix(s$RNA@scale.data[sigsub, ]))))

dav <- d3 %>% 
  group_by(celltype_label_3) %>% 
  summarize_at(.vars = sigsub, .funs = mean) %>% 
  `rownames<-`(value = NULL) %>%
  column_to_rownames("celltype_label_3")


cu <- diverge_hcl(palette = "Berlin", n = 12)
  
xy <- pheatmap(dav, 
                   color = cu,
                   scale = "column", 
                   width = 7,
                   height = 4, 
                   border_color = NA,
                   treeheight_col = 10, 
                   treeheight_row = 10,
                   silent = TRUE)

# extract celltype order 
celltype_order <- xy$tree_row$labels[xy$tree_row$order]
gene_order <- xy$tree_col$labels[xy$tree_col$order]

# summarize by percent expression 
gene_index <- sigsub
index1 <- gene_index[1]
index2 <- gene_index[length(gene_index)]

# calculate percent of cells with non zero expression 
pg0 <- function(x){
 perc(x = x,dir = "gt",val = 0)
}

# summary stat 
dsummary <- d3 %>% 
  group_by(celltype_label_3) %>% 
  summarize_at(.vars = sigsub, .funs = pg0) %>% 
  pivot_longer(cols = -celltype_label_3, names_to = "gene", values_to = "pct_express")

# calculate average normalized expression per cluster 
q90 <- function(x){
  quantile(x,probs = 0.95)
}
d_av <- 
  d3 %>% 
  group_by(celltype_label_3) %>% 
  summarize_at(.vars = sigsub, .funs = mean) %>% 
  pivot_longer(cols = -celltype_label_3, names_to = "gene", values_to = "gene_avg")

# merge summarized data 
d <- full_join(dsummary, d_av)

d$celltype_label_3 <- factor(d$celltype_label_3, levels = celltype_order)
d$gene <- factor(d$gene, levels = gene_order)
p <-
  ggplot(d %>% filter(pct_express > 0),
         aes(y = celltype_label_3, x = gene, 
             color = gene_avg,
             size = pct_express)) +
  geom_point() + 
  scale_color_gradientn(colours = viridis::viridis(n = 10),
                        limits = c(0,0.25), 
                        oob = scales::squish,
                        name = 'avg. expression') + 
  theme_bw() + 
  theme(axis.line  = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, color = 'black')) +
  theme(axis.text.y = element_text( color = 'black')) +
  ylab('') +
  theme(axis.ticks = element_blank()) 
print(p)
```
   
## Figure 6. Endotypes and microbiome
### Figure 6a. Boxplot showing the Bacterial/Viral metascore as a function of the pre-vaccination inflammatory endotypes. 
Read prevax cluster
```{r, read-slea, eval=TRUE}
inflamDF <- pData(sleaSet)[sleaSet$study_time_collected <= 0, ]
```
   
Bacterial.viral classifier
```{r, fig6, fig.height=4.4, fig.width=4.6, eval=TRUE}
# scale pre-vaccination
batch <- interaction(young.NoResp$study_accession,
                     young.NoResp$matrix)
scaledMat <- by(data    = t(exprs(young.NoResp)),
                INDICES = batch,
                FUN     = function(x) scale(x)) %>%
  do.call(what = rbind) %>%
  t()
scaledMat <- scaledMat[, young.NoResp$uid]

# higher in viral infections (IFI27, JUP, and LAX1) and
# higher in bacterial infections (HK3, TNIP1, GPAA1, and CTSB)
bvDF <- data.frame(score = colMeans(scaledMat[c("TNIP1", "GPAA1","CTSB"), ],
				    na.rm = TRUE) -
		     colMeans(scaledMat[c("IFI27", "JUP", "LAX1"), ],
				  na.rm = TRUE)) %>%
  rownames_to_column(var = "uid") %>%
  merge(y = inflamDF, by = "uid")

ggplot(data    = bvDF,
       mapping = aes(x = gr, y = score)) +
  geom_beeswarm(cex = 1.1, size = 0.75) +
  geom_boxplot(outlier.color = "transparent",
               fill          = "transparent",
               mapping       = aes(color = gr)) +
  labs(y = "Bacterial/Viral metascore", x = "Prevax inflammatory cluster") +
  theme_bw() +
  theme(legend.pos = "none")

kruskal.test(formula = score~gr, data = bvDF)
```
   
### Figure 6b. Gene expression of the inflammatory genes in dendritic cells stimulated with five PRR ligands
Read GSE134874 count file
```{r, read-counts, eval=TRUE}
countFile <- file.path(dataCacheDir, 
                       "GSE134874_Human_umi_counts.tsv.gz")
suppressWarnings(countDF <- read_tsv(file = countFile))

message("SampleName: Treatment_Donor_Replicate")
message(paste0("Treatment(n=8): Ctrl, G=cGAMP, H=polyIC, S=SeV, C=CpG-B, Z=Zymosan,",
               "L=LPS, P=Pam3CSK4"))
message(paste0("Ligand/Receptor: TLR4_L, TLR2_P, TLR3/MDA5_H, TLR9_C, RIG-I_S,",
               "Dectin-1_Z, STING-G"))
message("Donor (n=3): 27, 39, 45")
message("Replicate (n=3):1, 2, 3")
warning("no CpG or Pam3CSK4 in human DC experiment")
countDF <- select(countDF,
                  `X1`,
                  matches(match = "^Ctrl_"),
                  matches(match = "^cG_"),
                  matches(match = "^H_"),
                  matches(match = "^S_"),
                  matches(match = "^Z_"),
                  matches(match = "^L_"))
```
                
Read Lewis features that are inflammatory markers
```{r, read-lewis-genes, eval=TRUE}
featFile <- file.path(dataCacheDir,
                      "Top500_Var_VarImp.csv")
featDF <- read_csv(file = featFile, 
                   col_names = c("FEATURE", "IMPORTANCE"),
                   skip = 1) %>%
  filter(IMPORTANCE >= 0)

lewisLS <- featDF$FEATURE

# intersect them with inflammatory genes
lewisLS <- intersect(lewisLS,
                     unlist(all_genesets[c("HALLMARK_INFLAMMATORY_RESPONSE",
              "HALLMARK_COMPLEMENT",
              "HALLMARK_IL6_JAK_STAT3_SIGNALING",
              "HALLMARK_TNFA_SIGNALING_VIA_NFKB")]))
``` 

```{r, fig6b, warning=FALSE, fig.width=8, fig.height=8, eval=TRUE}
mat <- countDF %>%
    filter(`X1` %in% lewisLS) %>%
    column_to_rownames(var = "X1")
mat <- log2(mat + 0.25)
pdata <- colnames(mat) %>%
    strsplit(split = "_") %>%
    do.call(what = rbind) %>%
    as.data.frame() %>%
    setNames(nm = c("Treatment", "ParticipantID", "Replicate")) %>%
    mutate(TreatmentLong = recode(Treatment,
                                  "Ctrl" = "Ctrl", 
                                  "cG"   = "cGAMP", 
                                  "H"    = "polyIC", 
                                  "S"    = "SeV", 
                                  "Z"    = "Zymosan", 
                                  "L"    = "LPS"),
           TreatmentLong = factor(TreatmentLong,
                                  levels = unique(TreatmentLong)))           
rownames(pdata) <- colnames(mat)
matCtrl <- mat[, pdata$Treatment %in% "Ctrl"] 
muCtrl <- by(t(matCtrl), 
   INDICES = pdata[colnames(matCtrl), "ParticipantID"],
   FUN = colMeans) %>%
    do.call(what = cbind)
lfcMat <- mat - muCtrl[, pdata$ParticipantID] %>%
    as.matrix()
col_fun <- colorRamp2(colors = c("blue", "white", "red"), 
                    breaks = c(-5, 0, 5))
Heatmap(matrix = lfcMat, 
        col = col_fun,
        cluster_columns = FALSE,
        name = "log2FC",
        column_split = pdata$TreatmentLong)
```
