---
title: "PreVax Random Forest Training"
author: "Lewis Tomalin"
date: "2/10/2022"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(caret)
require(parallel)
require(limma)
require(plyr)
require(tidyverse)
require(tidytext)
require(caretEnsemble, lib.loc = "R")
require(caTools)

```

## Loading Pre-vaccination gene-expression data

The code below loads the gene-expression data, as well as the phenotype data (contains response information for each patient).

The phenotype data is filtered to include only high and low responders.

```{r data}

# Gene expression
load("ML.Data.BL.GENES.NoFilter_2021-06-14.rda")

# Phenotype 
load("PD.Master.BL.GENES.2021-06-14.rda")

PatientInfo %>% 
  filter(MFC_p30%in%c("lowResponder","highResponder")) %>%
  droplevels() %>%
  mutate(MFC_p30=fct_relevel(MFC_p30, c("highResponder", "lowResponder"))) -> PatientInfo

```

## Specifying X (input) and Y (response) data, and filtering

DevID contains PatientID for high and low responders.

The X and Y is subset to include only these patients, X.Dev Y.Dev (development data)

The development data is then filtered to include only the top 500 most varying genes.

```{r create training data, echo=FALSE}

DevID<-rownames(PatientInfo)
DevID<-intersect(DevID, colnames(data.ML))
X.Dev<-data.ML[,DevID]
rownames(X.Dev)<-strsplit2(rownames(X.Dev), "_")[,1]

Y.Dev<-PatientInfo[DevID, 'MFC_p30']
names(Y.Dev)<-DevID

## Keep top 500 most varying genes ##
VarVec<-sort(apply(X.Dev, 1, var), decreasing = T)
keep.genes<-names(VarVec[1:500])

X.Train.t1<-t(X.Dev[keep.genes,])

Y.Train1<-PatientInfo[rownames(X.Train.t1), 'MFC_p30']
names(Y.Train1)<-rownames(X.Train.t1)

```

## Define parameters for training using trainControl()

Ten fold-CV (adaptive), selecting the tuning parameters with highest AUROC. 


```{r}
SEED<-1987 # a great scientist was born this year

set.seed(SEED)
ctrl=trainControl(method = "adaptive_cv", 
                  numer=10,
                  adaptive = list(min = 5, alpha = 0.05,
                                  method = "gls", complete = TRUE),
                  selectionFunction = 'best',
                  search="grid",
                  returnResamp = "final",
                  classProbs = TRUE,
                  returnData = F,
                  verboseIter = F,
                  summaryFunction = twoClassSummary,
                  savePredictions = T)


```

## Run model training for random forest classifier

Training data is centered and scaled. 

A tune length of 10 is used. 

Model is trained to maximize the AUROC.

Parallelization code is provided for mac or windows.

```{r}

## Parallel on mac
# doMC::registerDoMC(detectCores()-2)

## Parallel on windows
#require(doParallel)
# cl<-makePSOCKcluster(detectCores()-2)
# registerDoParallel(cl)

set.seed(SEED)
model_list <- train(x=X.Train.t1,
                    y=Y.Train1[rownames(X.Train.t1)],
                    trControl=ctrl,
                    method="rf",
                    metric="ROC", 
                    tuneLength=10,
                    maximize=TRUE,
                    preProc = c("center", "scale"))

#stopCluster(cl)

#save(model_list, file="Top500_Var_Model.rda")


```

## Get Model Predictions and Variable Importance

Extracts 10-CV predictions, Training predictions and feature importance

```{r}

## Feature importance ##
db.imp<-arrange(varImp(model_list)$importance, desc(Overall))
#write.csv(db.imp, file="Top500_Var_VarImp.csv")

## Model Predictions 10-CV ##
db.preds<-model_list$pred %>%
  mutate(participant_id=rownames(X.Train.t1)[rowIndex],
         pathogen_vaccine_type=PatientInfo[rownames(X.Train.t1)[rowIndex],"pathogen_vaccine_type"])

preds.train <- predict(model_list, newdata=X.Train.t1, type="prob")
preds.train <- data.frame(preds.train) %>% mutate(obs=Y.Train1[rownames(preds.train)])

save(db.preds, preds.train, file="DB.PREDS.rda")

```



















